---
title: "Data Cleaning Group Project"
author: "Melisa Abigail Thevandi"
date: "2024-10-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, 
                      include = TRUE,
                      fig.align = "center",  out.width = "80%")
```

```{r}
# Load necessary packages
library(lubridate)
library(readxl)
library(tidyverse)
library(stringr)
# Disable scientific notation
options(scipen = 999)
```

```{r}
# install.packages("tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load('2023-10-17')

taylor_album_songs <- tuesdata$taylor_album_songs
taylor_all_songs <- tuesdata$taylor_all_songs
taylor_albums <- tuesdata$taylor_albums
```

# Missing Values is not being removed since it indicates that there is no data.

# Check the uniqueness

### taylor_album_songs
```{r}
glimpse(taylor_album_songs)
```
##### Number of album
```{r}
taylor_album_songs %>% count(album_name)
```
#### Filter track_name that has n > 1
```{r}
taylor_album_songs %>% count(track_name) %>% filter(n > 1)
# None
```


### taylor_all_songs
```{r}
glimpse(taylor_all_songs)
```
#### Number of album_name occurrences
```{r}
taylor_all_songs %>% count(album_name)
# the album that's present in taylor_album_songs has the number of occurrences
# some albums are not present in taylor_album_songs
```
#### Checking the albums in taylor_all_songs that are not present in taylor_album_songs
```{r}
anti_join(taylor_all_songs, taylor_album_songs, by = "album_name")
```
```{r}
anti_join(taylor_all_songs, taylor_album_songs, by = "album_name") %>% count(album_name)
# album_name Beautiful Eyes, Fearless, Red, and The Taylor Swift Holiday Collection are not
# in taylor_album_songs;
# 
# songs with album_name=NA means that the songs are not in any album;
#
# songs in Beautiful Eyes have a lot of missing data
```
#### Filter track_name that has n > 1
```{r}
taylor_all_songs %>% count(track_name) %>% filter(n > 1)
# I'm Only Me When I'm With You appears twice
```
#### Check the duplicate track_name
```{r}
taylor_all_songs %>% filter(track_name %in% "I'm Only Me When I'm With You")
# the song appears in Taylor Swift and Beautiful Eyes album
```

### taylor_albums
```{r}
glimpse(taylor_albums)
```
#### Filter album_name that has n > 1
```{r}
taylor_albums %>% count(album_name) %>% filter(n > 1)
# None
```



# Handling Some Missing Data

### Change artist = NA to artist = "Taylor Swift" in taylor_all_songs, for album_name = Beautiful Eyes
```{r}
taylor_all_songs$artist <- ifelse(taylor_all_songs$album_name == "Beautiful Eyes",
                                  "Taylor Swift", taylor_all_songs$artist)
```
```{r}
taylor_all_songs
```


# Joining Data

### Joining by album for taylor_album_songs and taylor_albums
```{r}
album <- left_join(taylor_album_songs, taylor_albums, by = "album_name") %>%
  select(album_name, ep = ep.x, album_release = album_release.x, track_number, track_name,
         artist, featuring, bonus_track, single_release, track_release, danceability,
         energy, key, loudness, mode, speechiness, acousticness, instrumentalness,
         liveness, valence, tempo, time_signature, duration_ms, explicit, key_name,
         mode_name, key_mode, metacritic_score, user_score)
glimpse(album)
# all rows have metacritic_score and user_score
```


```{r}
write.csv(album, "../data/album.csv", row.names=FALSE)
```

