---
title: "DSA2101 Group Project"
author: ""
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, include = TRUE,
                      fig.align = "center",  out.width = "80%")
library(tidyverse)
library(stringr)
library(lubridate)
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(viridis)
library(gridExtra) 
library(reshape2)
library(gganimate)
library(glue)
# Disable scientific notation
options(scipen = 999)
```

```{r}
# install.packages("tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load('2023-10-17')

taylor_album_songs <- tuesdata$taylor_album_songs
taylor_all_songs <- tuesdata$taylor_all_songs
taylor_albums <- tuesdata$taylor_albums
```

# Missing Values is not being removed since it indicates that there is no data.

# Check the uniqueness

### taylor_album_songs
```{r}
glimpse(taylor_album_songs)
```
##### Number of album
```{r}
taylor_album_songs %>% count(album_name)
```
#### Filter track_name that has n > 1
```{r}
taylor_album_songs %>% count(track_name) %>% filter(n > 1)
# None
```


### taylor_all_songs
```{r}
glimpse(taylor_all_songs)
```
#### Number of album_name occurrences
```{r}
taylor_all_songs %>% count(album_name)
# the album that's present in taylor_album_songs has the number of occurrences
# some albums are not present in taylor_album_songs
```
#### Checking the albums in taylor_all_songs that are not present in taylor_album_songs
```{r}
anti_join(taylor_all_songs, taylor_album_songs, by = "album_name")
```
```{r}
anti_join(taylor_all_songs, taylor_album_songs, by = "album_name") %>% count(album_name)
# album_name Beautiful Eyes, Fearless, Red, and The Taylor Swift Holiday Collection are not
# in taylor_album_songs;
# 
# songs with album_name=NA means that the songs are not in any album;
#
# songs in Beautiful Eyes have a lot of missing data
```
#### Filter track_name that has n > 1
```{r}
taylor_all_songs %>% count(track_name) %>% filter(n > 1)
# I'm Only Me When I'm With You appears twice
```
#### Check the duplicate track_name
```{r}
taylor_all_songs %>% filter(track_name %in% "I'm Only Me When I'm With You")
# the song appears in Taylor Swift and Beautiful Eyes album
```

### taylor_albums
```{r}
glimpse(taylor_albums)
```
#### Filter album_name that has n > 1
```{r}
taylor_albums %>% count(album_name) %>% filter(n > 1)
# None
```



# Handling Some Missing Data

### Change artist = NA to artist = "Taylor Swift" in taylor_all_songs, for album_name = Beautiful Eyes
```{r}
taylor_all_songs$artist <- ifelse(taylor_all_songs$album_name == "Beautiful Eyes",
                                  "Taylor Swift", taylor_all_songs$artist)
```
```{r}
taylor_all_songs
```


# Joining Data

### Joining by album for taylor_album_songs and taylor_albums
```{r}
album <- left_join(taylor_album_songs, taylor_albums, by = "album_name") %>%
  select(album_name, ep = ep.x, album_release = album_release.x, track_number, track_name,
         artist, featuring, bonus_track, single_release, track_release, danceability,
         energy, key, loudness, mode, speechiness, acousticness, instrumentalness,
         liveness, valence, tempo, time_signature, duration_ms, explicit, key_name,
         mode_name, key_mode, metacritic_score, user_score)
glimpse(album)
# all rows have metacritic_score and user_score
```





```{r}
# Convert track_release to Date format and extract year
album$track_release <- as.Date(album$track_release)
album$year_group <- cut(album$track_release, 
                        breaks = as.Date(c("2005-01-01", "2010-01-01", "2015-01-01", "2020-01-01", "2025-01-01")),
                        labels = c("2006-2009", "2010-2014", "2015-2019", "2020-2024"))

# Function to calculate correlation coefficient for each year group
get_correlation <- function(album, x, y) {
  album %>%
    group_by(year_group) %>%
    summarize(correlation = round(cor(.data[[x]], .data[[y]], use = "complete.obs"), 2))
}

# Calculate correlation coefficients for energy vs loudness
correlations <- get_correlation(album, "loudness", "energy")

# Merge correlations back into the original data
album <- merge(album, correlations, by = "year_group")

# Create scatter plot with LOESS lines and 95% confidence intervals for energy vs loudness
ggplot(album, aes(x = loudness, y = energy)) +
  geom_point() +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  facet_wrap(~ year_group) +
  
  # Add correlation coefficient as a text annotation for each year group
  geom_text(data = correlations, aes(label = sprintf("r = %.2f", correlation), x = -12, y = 0.9), 
            inherit.aes = FALSE, size = 3.7, color = "blue") + 
  labs(title = "Energy vs Loudness by Year Group",
       subtitle = "r represents the correlation coefficient between energy and loudness",
       x = "Loudness",
       y = "Energy",
       caption = "The grey band represents the 95% confidence interval around the LOESS trendline.") +
  theme_minimal()

```


```{r}

# Calculate the percentage of songs by key
key_distribution <- album %>%
  group_by(key_name, album_release) %>%
  summarize(count = n()) %>%
  mutate(percentage = (count / sum(count, na.rm = T)) * 100) %>%
  na.omit()

# Define the positions for white and black keys on a piano
white_keys <- c("C", "D", "E", "F", "G", "A", "B")
black_keys <- c("C#", "D#", "F#", "G#", "A#")

# Define positions for keys on the keyboard (x-axis)
key_positions <- data.frame(
  key_name = c(white_keys, black_keys),
  position = c(1, 2, 3, 4, 5, 6, 7, 1.5, 2.5, 4.5, 5.5, 6.5) # Black keys centered between white keys
)

dfcur <- NULL
for (y in sort(unique(key_distribution$album_release))){
  temp = key_distribution %>%
    filter(album_release == y) %>%
    full_join(key_positions,  by = "key_name") %>%
    replace_na(list(count = 0, percentage = 0))
  temp$album_release = y
  dfcur <- rbind(dfcur, temp)
}

key_distribution <- dfcur

# Create a custom keyboard plot
animated_plot <- ggplot() +
  
  # White keys (rectangles side by side with no gaps)
  geom_rect(data = key_distribution %>% filter(key_name %in% white_keys),
            aes(xmin = position - 0.5, xmax = position + 0.5,
                ymin = -0.3, ymax = 0.8), fill = "#FFFFFF", color = "#000000") +
  
  # Black keys (rectangles centered on top of white keys)
  geom_rect(data = key_distribution %>% filter(key_name %in% black_keys),
            aes(xmin = position - 0.25, xmax = position + 0.25,
                ymin = 0.1, ymax = 0.8), fill = "#000000") +
  
  # Add text labels for white keys with gradient color based on percentage and white text
  geom_label(data = key_distribution %>% filter(key_name %in% white_keys),
             aes(x = position, y = -0.5,
                 label = key_name,
                 fill = percentage), color = "#FFFFFF", size = 6.7) + # White text for contrast
  
  # Add text labels for black keys with gradient color based on percentage and black text
  geom_label(data = key_distribution %>% filter(key_name %in% black_keys),
             aes(x = position, y = 0.325,
                 label = key_name,
                 fill = percentage), color = "#000000", size = 6) + # Black text for contrast
  
  # Use viridis color scale for percentage gradient
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  
  # Adjust plot limits and remove axis labels/ticks to resemble a keyboard
  scale_y_continuous(limits = c(-2, NA)) +
  
  labs(title = "{closest_state} - Percentage of Songs by Key",
       x = NULL,
       y = NULL,
       fill = "% of Songs") +
  
  theme_void() +   # Remove axis lines and ticks
  theme(legend.position="top") +
  
  # Animate over album_release
  transition_states(album_release) +
  enter_fade() + exit_fly(y_loc = 1)

# Save the animation as a GIF
animate(animated_plot, duration = 10, fps = 1, renderer = av_renderer())

```

